---
title: "Calculate drop in body weight"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculate drop in body weight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
This vignette assumes a SQL server at `localhost` (we use PostgreSQL), with an OMOP Common Data Model v5.4 in schema `cdm_new_york3` containing patient records. The patient records shown in this example are synthetic data from [Synthea Patient Generator](https://github.com/synthetichealth/synthea).  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# kury - credenciais.R contains the cred$pg credentials.
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/espinha/credenciais.R'))
```
```{r setup}
library(knitr)
library(phea)
suppressPackageStartupMessages(library(dplyr))

# Connect to SQL server.
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
    host = 'localhost', port = 7654,
    dbname = 'fort', user = cred$pg$user, password = cred$pg$pass)

# Provide the connection to Phea so we can use the sqlt() and sql0() shorthands.
setup_phea(dbcon, 'cdm_new_york3')
```

Say we want to find patients whose weight has increased by more than 50% in the last 2 years to 3 years. The formula for the number we want is very simple:  

`weight_drop = (weight_current - weight_previous) / weight_previous`  

If `weight_drop` is ever ≥0.5, that means an increase of ≥50%. But unlike in other formulas, here the component `weight_previous` is not the most recently available weight record for the given patient, at the given point in time.  

While the _record source_ is the same as `weight_current`, the _component_ is different. When we call `make_component()`, we need to specify the time frame using the `delay` option (and optionally the `window` as well).  

```{r delay}
# Weight records
# Loinc 29463-7 Body weight, concept ID 3025315
weight_record_source = sqlt(measurement) |>
  filter(measurement_concept_id == 3025315) |>
  make_record_source(
    ts = measurement_datetime,
    pid = person_id)

# Most recent weight count record
weight_current = make_component(weight_record_source)

# weight count record at least 24 hours older than phenotype date, up to limit of 48 hours
weight_previous = make_component(weight_record_source,
  delay = '2 year',
  window = '4 years')
```

_`delay` picks the most recent record that is **at least as old** as the specified amount of time._  

_`window` **limits how old** the record is allowed to be._  

Both are expressed in SQL language, including the automatic parsing, by the server, of terms such as `'2 years'`, `'2.5 hours'`, `'1 week'`, etc.  

If no such record exists for the patient, at the given point in time, the value is `NULL`.  

## Calculate the phenotype  

Armed with the two components, we call `calculate_formula()`:  

```{r phenotype}
phen <- calculate_formula(
  components = list(
    weight_current = weight_current,
    weight_previous = weight_previous),
  fml = list(
    weight_change = '(weight_current_value_as_number - weight_previous_value_as_number) /
      weight_previous_value_as_number',
    weight_50p_increase = "case when weight_change >= 0.5 then 'yes' else 'no' end"),
  export = list(
    'weight_current_measurement_datetime',
    'weight_previous_measurement_datetime')) |>
  filter(!is.na(weight_change))
```

## Plot the phenotype result  
```{r, include = FALSE}
# # Source:     SQL [?? x 3]
# # Database:   postgres  [postgres@localhost:7654/fort]
# # Ordered by: desc(uvals)
#      pid   lines   uvals
#    <int> <int64> <int64>
#  1 10473      78      78
#  2  6028      75      75
#  3 11176      74      74
#  4  5601      75      74
#  5 21264      73      73
#  6  4172      73      72
#  7 14117      72      71
#  8 20051      71      71
#  9 10617      71      71
# 10 19133      72      70

# person_id # linhas # dists # 1
# 11872 # 133 # 81 # 2
# 22242 # 158 # 78 # 3
# 22906 # 182 # 76 # 4
# 4191 # 141 # 75 # 5
# 20198 # 160 # 73 # 6
# 5274 # 136 # 72 # 7
# 18843 # 132 # 71 # 8
# 1665 # 162 # 69 # 9
# 10583 # 125 # 69 # 10
# 5333 # 170 # 68
```

## Filter patients whose increase is at least 50%  
The `phen` object is a _lazy table_, which means it is a SQL query. We can manipulate it to obtain only the patients that had a ≥50% increase in body weight at least once in their history.  

```{r random_patient_plot, out.width="100%", out.height = 500}
patients_50p <- phen |>
  filter(weight_change >= 0.5) |> # Only rows where increase is ≥50%
  group_by(pid) |> # Group by patient ID
  summarise(
    first_50p_increase = min(ts, na.rm = TRUE)) |> # Keep timestamp of earliest increase of ≥50% for each patient
  collect() # Execute query and download data from the server to local memory

# Sample one patient at random, from among those with ≥50% weight increase at least once in their history.
sample_patient <- sample(patients_50p$pid, 1)
message('Sampled patient: ', sample_patient)

# Plot the phenotype for the sampled patient.
phen |>
  select(
    -weight_current_measurement_datetime,
    -weight_previous_measurement_datetime,
    -weight_50p_increase) |>
  phea_plot(pid = sample_patient)
```

## Obtain the SQL query that computes the phenotype  
To see the SQL query underlying the phenotype, use helper function `code_shot()`, or `dbplyr::sql_render()`, or the `.clip_sql` option in `calculate_formula()`.  
```{r, eval = FALSE}
code_shot(phen)
```
```sql
`r code_shot(phen)`
```

```{r, include = FALSE}
DBI::dbDisconnect(dbcon)
```
