---
title: "Kidney injury lab presentation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acute kidney injury labs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  _This vignette assumes a SQL server at `localhost` (we use PostgreSQL), with data in OMOP Common Data Model v5.4 format in schema `cdm_new_york3`. The patient records shown in this example are synthetic data from [Synthea(TM) Patient Generator](https://github.com/synthetichealth/synthea)._  
```{r, results = 'hide', include = FALSE}
library(knitr)
options(scipen = 5e5)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/espinha/credenciais.R')) # contains cred$pg
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/wrap/wrap.R')) # contains cred$pg
```
```{r setup, message = FALSE}
library(phea)
library(dplyr)

# Connect to SQL server.
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
  host = 'localhost', port = 7654, dbname = 'fort',
  user = cred$pg$user, password = cred$pg$pass)

# Call setup_phea so we can use sqlt() and sql0().
setup_phea(dbcon, 'cdm_new_york3')
```

In this vignette we identify:  

Case A.  

 - Increase in serum creatinine (SCr) by ≥0.3 mg/dL (≥26.5 µmol/L) within 48 hours;  
 
Case B.

 - Increase in serum creatinine (SCr) to ≥1.5 times baseline within 7 days;  
 
Case C.

 - Two GFR measurements, at least 3 but less than 5 months apart, showing value < 60 ml/min/1.73m².

Here's how we compute it:  

 - In a., for every measurement of SCr, we check if we can find another one, within the prior 48 hours, where the value is smaller by a margin ≥ 0.3 mg/dL.
 - In b. it is similar, but with a 7-day window instead of 48 hours, and the check is whether the values exhibit a ratio of ≥ 1.5, instead of a difference of ≥ 0.3.  
 - In c., for every measurement of GFR that is < 60, we look to see if there is one more inside the 3-to-5 months window.  

In both cases, we will normalize the unit of measurement of serum creatinine to md/dL prior to computing formulas. We could also use formulas for that, and the result would be the same.  

## Create components  
### Serum creatinine  
We collect SCr records from `MEASUREMENT` and convert the units to mg/dL.  
```{r scr_records}
# Serum creatinine codes used: 
# Loinc 38483-4 Creatinine [Mass/volume] in Blood, OMOP CDM concept ID 3051825
# Loinc 2160-0 Creatinine [Mass/volume] in Serum or Plasma, OMOP CDM concept ID 3016723

# "A" records: Unit is 'mg/dL'.
scr_records_a <- sqlt(measurement) |>
  filter(measurement_concept_id %in% c(3051825, 3016723) &&
    unit_source_value == 'mg/dL')

# "B" records: Unit is 'µmol/L', but we convert to 'mg/dL'.
scr_records_b <- sqlt(measurement) |>
  filter(measurement_concept_id %in% c(3051825, 3016723) &&
    unit_source_value == 'µmol/L') |>
  mutate( # Convert µmol/L to mg/dL
    value_as_number = value_as_number / 88.42,
    unit_source_value = 'mg/dL')

# Combine all available serum creatinine records.
scr_records <- union_all(scr_records_a, scr_records_b)

# Make a record source.
scr_record_source <- make_record_source(scr_records,
  ts = measurement_datetime,
  pid = person_id)
```

### Glomerular filtration rate  
We collect GFR records from `MEASUREMENT`.
```{r gfr_records}
# GFR codes used:
# Loinc 77147-7 Glomerular filtration rate/1.73 sq M.predicted [Volume Rate/Area] in Serum, Plasma or Blood by Creatinine-based formula (MDRD)
# OMOP CDM concept ID 46236952 
gfr_records <- sqlt(measurement) |>
  filter(measurement_concept_id == 46236952)

gfr_record_source <- make_record_source(gfr_records,
    ts = measurement_datetime,
    pid = person_id)
```

## Calculate the phenotype  
Formula `scr_case_a` contains the logic of case A (difference ≥0.3 mg/dL), `scr_case_b` contains case B (ratio ≥1.5), and `gfr_case_c` contains case C (two GFRs under 60 within 3-5 months).  
  
**Notice we use 48 days instead of 48 hours, and 7 months instead of 7 days. This is just to make a visually meaningful example out of the limited data that is produced by Synthea(TM). Those time intervals can be changed to "48 hours" and "7 days" to produce the real phenotype.**  
```{r phenotype}
scr_change <- calculate_formula(
  components = list(
    # Current SCr
    scr = make_component(scr_record_source),
    
    # Minimum value within 48-hour window
    scr_48h_min = make_component(scr_record_source,
      window = '48 days', .delay_fn = 'min'),
    
    # Minimum value within 7-day window
    scr_7d_min = make_component(scr_record_source,
      window = '7 months', .delay_fn = 'min'),
    
    # Current glomerular filtration rate (GFR)
    gfr = make_component(gfr_record_source),
    
    # Glomerular filtration rate 3 to 5 months older than phenotype date
    gfr_prior = make_component(gfr_record_source,
      delay = '3 months', window = '5 months', .delay_fn = 'min')),
    
  fml = list(
    scr_case_a = 'scr_value_as_number - scr_48h_min_value_as_number >= 0.3',
    
    scr_case_b = 'scr_value_as_number / scr_7d_min_value_as_number >= 1.5',
    
    gfr_case_c = 'gfr_value_as_number < 60 AND gfr_prior_value_as_number < 60'),
  
  export = c(
    'scr_measurement_datetime',
    'scr_48h_min_measurement_datetime',
    'scr_7d_min_measurement_datetime',
    'gfr_measurement_datetime',
    'gfr_prior_measurement_datetime'),
  
  .cascaded = FALSE, # Because we don't need to use results of prior formulas inside other formulas.
)

```

Let us take a small peek at 15 rows from the phenotype results.  
```{r}
head_shot(scr_change, 15) |>
  kable()
```

## Plot the phenotype for a random patient  
For demonstration purposes, let us pick an "interesting" patient. By "interesting" I just mean a patient who at different times was `TRUE` and `FALSE` in each of the three criteria. A patient with variability in their timeline.   
```{r filter}
# Find an "interesting" case to plot: patients with each criteria at times TRUE, at times FALSE
patients <- scr_change |>
  group_by(pid) |>
  summarise(
    n_03 = n_distinct(scr_case_a),
    n_15 = n_distinct(scr_case_b),
    n_gfr = n_distinct(gfr_case_c)) |>
  mutate(sort_variable = n_03 + n_15 + n_gfr) |>
  arrange(desc(sort_variable)) |>
  head(20) |>
  select(pid) |>
  pull()

random_patient <- sample(patients, 1)

message('Sampled patient: ', random_patient)
```

Then we plot all data for the chosen patient (pid = ``r random_patient``).  
```{r plot, out.width="100%", out.height = 1400}
scr_change |>
  select(-ends_with('datetime')) |>
  mutate(
    scr_case_a_or_b = scr_case_a | scr_case_b) |>
  phea_plot(random_patient)
```

_At the end of this report I include a large table with all the data for this patient, directly from the record sources, for maximum verification._  

## Obtain the SQL query that computes the phenotype  
To see the SQL query underlying the phenotype, use helper function `code_shot()`, or `dbplyr::sql_render()`, or the `.clip_sql` option in `calculate_formula()`.  
```{r, eval = FALSE}
code_shot(scr_change)
```
```sql
`r code_shot(scr_change)`
```

## See the raw data that went into `calculate_formula()` 
Sometimes the best way to check a result is to manually, painstakingly go over the data. Below I print all rows from the record sources for the patient that was plotted above.  

#### Serum creatinine records  
```{r plot_data_scr}
scr_records |>
  filter(person_id == random_patient) |>
  collect() |>
  arrange(measurement_datetime) |>
  select(measurement_id, person_id, measurement_datetime, value_as_number, unit_source_value) |>
  kable()
```

#### Glomerular filtration rate records
```{r plot_data_gfr}
gfr_records |>
  filter(person_id == random_patient) |>
  collect() |>
  arrange(measurement_datetime) |>
  select(measurement_id, person_id, measurement_datetime, value_as_number, unit_source_value) |>
  kable()
```

## Produce events for Atlas  
One approach to use Phea's results inside Atlas is to produce novel records (e.g. new rows in `OBSERVATION` or `MEASUREMENT`) using custom concept IDs, then ETL those back into the dataset.  

Below I illustrate how to do it with case A, assuming the new records will go into `MEASUREMENT`. Notice I put the difference inside `value_as_number`, just in case.    
```{r change_events}
case_a_custom_concept_id <- 2000000001
# Case A: SCr increase by >=0.3 over 48 hours
data_for_etl <- scr_change |>
  filter(scr_case_a) |> # Keep only rows where case A was TRUE
  transmute(
    measurement_datetime = ts,
    person_id = pid,
    value_as_number = scr_value_as_number - scr_48h_min_value_as_number,
    measurement_concept_id = case_a_custom_concept_id)

head_shot(data_for_etl) |>
  kable()
```

### Author contact  
Fabrício Kury -- Please be always welcome to reach me at fab@kury.dev.  

```{r, include = FALSE}
DBI::dbDisconnect(dbcon)
```
