---
title: "Acute kidney injury lab presentation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acute kidney injury labs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  _This vignette assumes a SQL server at `localhost` (we use PostgreSQL), with data in OMOP Common Data Model v5.4 format in schema `cdm_new_york3`. The patient records shown in this example are synthetic data from [Synthea(TM) Patient Generator](https://github.com/synthetichealth/synthea)._  
```{r, results = 'hide', include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/espinha/credenciais.R')) # contains cred$pg
```
```{r setup, print = FALSE}
library(phea)
library(dplyr)

# Connect to SQL server.
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
  host = 'localhost', port = 7654, dbname = 'fort',
  user = cred$pg$user, password = cred$pg$pass)

# Call setup_phea so we can use sqlt() and sql0().
setup_phea(dbcon, 'cdm_new_york3')
```

In this vignette we identify:  

Case A.  

 - Increase in serum creatinine (SCr) by ≥0.3 mg/dL (≥26.5 µmol/L) within 48 hours;  
 
Case B.

 - Increase in serum creatinine (SCr) to ≥1.5 times baseline within 7 days;  

Here's how we compute it:  

 - In a., for every measurement of SCr, we check if we can find another one, within in the prior 48 hours, where the value is smaller by a margin >= 0.3 mg/dL.
 - In b. it is similar, but with a 7-day window instead of 48 hours, and the check is whether the values exhibit a ratio of >= 1.5, instead of a difference of >= 0.3.  

In both cases, we will normalize the unit of measurement of SCr to md/dL prior to computing formulas. We could also use formulas to convert the units, and the result would be the same.  

## Create components  
### Serum creatinine  
We collect SCr records from `MEASUREMENT` and convert the units to mg/dL.  
```{r scr_records}
# 3051825 38483-4 Creatinine [Mass/volume] in Blood
# 3016723 2160-0 Creatinine [Mass/volume] in Serum or Plasma

# "A" records: Unit is 'mg/dL'
scr_records_a <- sqlt(measurement) |>
  filter(measurement_concept_id %in% c(3051825, 3016723) &&
    unit_source_value == 'mg/dL')

# "B" records: Unit is 'µmol/L', but we convert to 'mg/dL'
scr_records_b <- sqlt(measurement) |>
  filter(measurement_concept_id %in% c(3051825, 3016723) &&
    unit_source_value == 'µmol/L') |>
  mutate( # Convert µmol/L to mg/dL
    value_as_number = value_as_number / 88.42,
    unit_source_value = 'mg/dL')

# Combine all available serum creatinine records into a single record source.
scr_record_source <- union_all(scr_records_a, scr_records_b) |>
  make_record_source(
    ts = measurement_datetime,
    pid = person_id)
```

### Glomerular filtration rate  
We collect GFR records from `MEASUREMENT`.
```{r gfr_records}
## GFR
# 46236952 77147-7 Glomerular filtration rate/1.73 sq M.predicted [Volume Rate/Area] in Serum, Plasma or Blood by Creatinine-based formula (MDRD)
gfr_record_source <- sqlt(measurement) |>
  filter(measurement_concept_id == 46236952) |>
  make_record_source(
    ts = measurement_datetime,
    pid = person_id)
```

## Calculate the phenotype  
Formula `scr_03_criteria` contains the logic of case A (difference >= 0.3 mg/dL), while `scr_15_criteria` contains case B (ratio >= 1.5).  
```{r phenotype}
scr_change <- calculate_formula(
  components = list(
    # Current SCr
    scr = make_component(scr_record_source),
    
    # Minimum value within 48-hour window
    scr_48h_min = make_component(scr_record_source,
      window = '48 days', .delay_fn = 'min'),
    
    # Minimum value within 7-day window
    scr_7d_min = make_component(scr_record_source,
      window = '7 months', .delay_fn = 'min'),
    
    # Current glomerular filtration rate (GFR)
    gfr = make_component(gfr_record_source),
    
    # Glomerular filtration rate 3 to 5 months older than phenotype date
    gfr_3mo = make_component(gfr_record_source,
      delay = '3 months', window = '5 months')),
    
  fml = list(
    scr_03_criteria = 'scr_value_as_number - scr_48h_min_value_as_number >= 0.3',
    
    scr_15_criteria = 'scr_value_as_number / scr_7d_min_value_as_number >= 1.5',
    
    gfr_criteria = 'gfr_value_as_number < 60 AND gfr_3mo_value_as_number < 60'),
  
  export = c(
    'scr_measurement_datetime',
    'scr_48h_min_measurement_datetime',
    'scr_7d_min_measurement_datetime',
    'gfr_measurement_datetime',
    'gfr_3mo_measurement_datetime')
)

```
```{r, eval = FALSE, include = FALSE}
head_shot(scr_change, 100)
```

```{r filter ami, eval = FALSE, include = FALSE}
# Keep only patients who had meaningful change
ami_patients <- ami_obese |>
  filter(has_ami) |>
  select(pid) |>
  distinct() |>
  pull()

ami_obese <- ami_obese |>
  filter(pid %in% local(ami_patients))

head_shot(ami_obese) |>
  kable()
```

## Plot the phenotype for a random patient
```{r, eval = FALSE, include = FALSE}
ami_obese_patients <- ami_obese |>
  filter(ami_obese) |>
  select(pid) |>
  distinct() |>
  pull()

random_patient <- sample(ami_obese_patients, 1)
message('Sampled patient: ', random_patient)
```
```{r plot, out.width="100%", out.height = 1200}
scr_change |>
  select(-ends_with('datetime')) |>
  phea_plot(22444)
```

## Obtain the SQL query that computes the phenotype  
To see the SQL query underlying the phenotype, use helper function `code_shot()`, or `dbplyr::sql_render()`, or the `.clip_sql` option in `calculate_formula()`.  
```{r, eval = FALSE, include = FALSE}
code_shot(ami_obese)
```

```{r, include = FALSE}
DBI::dbDisconnect(dbcon)
```
