---
title: "Calculating the ASCVD Risk Estimator Plus score"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASCVD+}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  _This vignette assumes a SQL server at `localhost` (we use PostgreSQL), with data in OMOP Common Data Model v5.4 format in schema `cdm_new_york3`. The patient records shown in this example are synthetic data from [Synthea(TM) Patient Generator](https://github.com/synthetichealth/synthea)._  
```{r, results = 'hide', include = FALSE}
library(knitr)
options(scipen = 5e5)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/espinha/credenciais.R')) # contains cred$pg
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/wrap/wrap.R'))

usn <- function(n, decimals = 0) {
  sapply(n, function(e) {
    if(is.na(e))
      return(e)
    # usn()'s decimals behave like round()'s digits, but comma()'s precision is different.
    scales::comma(e, 10^(-decimals))
  })
}
```
```{r setup, message = FALSE}
library(phea)
library(dplyr)

# Connect to SQL server.
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
  host = 'localhost', port = 7654, dbname = 'fort',
  user = cred$pg$user, password = cred$pg$pass)

# Call setup_phea so we can use sqlt() and sql0().
setup_phea(dbcon, 'cdm_new_york3')
```

## Components

### Systolic Blood Pressure (mm Hg)
3004249 Loinc 8480-6 Systolic blood pressure
```{r systolic_blood_pressure}
sbp <- sqlt(measurement) |>
  filter(measurement_concept_id == 3004249) |>
  make_component(
    .pid = person_id,
    .ts = measurement_datetime)
```

### Diastolic Blood Pressure (mm Hg)
3012888 Loinc 8462-4 Diastolic blood pressure
```{r diastolic_blood_pressure}
dbp <- sqlt(measurement) |>
  filter(measurement_concept_id == 3012888) |>
  make_component(
    .pid = person_id,
    .ts = measurement_datetime)
```

### Total Cholesterol (mg/dL)
3027114 Loinc 2093-3 Cholesterol [Mass/volume] in Serum or Plasma
```{r total_cholesterol}
tc <- sqlt(measurement) |>
  filter(measurement_concept_id == 3027114) |>
  make_component(
    .pid = person_id,
    .ts = measurement_datetime)
```

### HDL Cholesterol (mg/dL)
3007070 Loinc 2085-9 Cholesterol in HDL [Mass/volume] in Serum or Plasma
```{r hdl_cholesterol}
hdl <- sqlt(measurement) |>
  filter(measurement_concept_id == 3007070) |>
  make_component(
    .pid = person_id,
    .ts = measurement_datetime)
```

### LDL Cholesterol (mg/dL)
3009966 Loinc 18262-6 Cholesterol in LDL [Mass/volume] in Serum or Plasma by Direct assay
```{r ldl_cholesterol}
ldl <- sqlt(measurement) |>
  filter(measurement_concept_id == 3009966) |>
  make_component(
    .pid = person_id,
    .ts = measurement_datetime)
```

### History of Diabetes?
201826 SNOMED 44054006 Type 2 diabetes mellitus
```{r diabetes}
diabetes <- sqlt(condition_occurrence) |>
  filter(condition_concept_id == 201826) |>
  make_component(
    .pid = person_id,
    .ts = condition_start_datetime)
```

#### Smoker?
_43054909 LOINC 72166-2 Tobacco smoking status_
42709996 SNOMED 449868002 Smokes tobacco daily
```{r smoker}
smoker <- sqlt(observation) |>
  filter(observation_concept_id %in% c(42709996)) |>
  make_component(
    .pid = person_id,
    .ts = observation_datetime)
```

### On Hypertension Treatment?
ATC:
	C02 ANTIHYPERTENSIVES
	C03 - Diuretics
	C07 - Beta blocking agents
	C08 - Calcium channel blockers and
	C09 - Agents acting on the renin-angiotensin system.
```{r hypertension_treatment}
atc_concepts <- sqlt(concept) |>
  filter(concept_code %in% c('C02', 'C03', 'C07', 'C08', 'C09') &&
      vocabulary_id == 'ATC') |>
  select(concept_id)

drug_concepts <- sqlt(concept_ancestor) |>
  inner_join(atc_concepts, by = c('ancestor_concept_id' = 'concept_id')) |>
  select(descendant_concept_id)

htt <- sqlt(drug_exposure) |>
  inner_join(drug_concepts, by = c('drug_concept_id' = 'descendant_concept_id')) |>
  make_component(
    .pid = person_id,
    .ts = drug_exposure_start_datetime)
```

#### _On a Statin?_  

#### _On Aspirin Therapy?_  

#### Patient age  
```{r person}
person <- sqlt(person) |>
  make_component(
    .pid = person_id,
    .ts = birth_datetime)
```

## Formula
```{r}
ascvd = calculate_formula(
  components = list(
    sbp = sbp,
    dbp = dbp,
    tc = tc,
    hdl = hdl,
    ldl = ldl,
    diabetes = diabetes,
    smoker = smoker,
    htt = htt,
    person = person),
  export = c(
    'sbp_value_as_number',
    'dbp_value_as_number',
    'tc_value_as_number',
    'hdl_value_as_number',
    'ldl_value_as_number',
    'diabetes_condition_concept_id',
    'diabetes_condition_start_datetime',
    'smoker_observation_datetime',
    'htt_drug_concept_id',
    'htt_drug_exposure_start_datetime',
    'person_birth_datetime'),
  fml = c(
    age = 'extract(year from age(ts, person_birth_datetime))',
    smoker = 'smoker_observation_datetime is not null',
    diabetes = 'diabetes_condition_start_datetime is not null',
    htt = "htt_drug_exposure_start_datetime is not null
      and abs(htt_drug_exposure_start_datetime - sbp_ts) < '4 months'", # SBP and HTT within 4 months of each other
    is_woman = 'person_gender_concept_id == 8532',
    is_black = 'person_race_concept_id == 8516'
    ),
  .dont_require = c(
    'smoker', 'htt', 'diabetes'))


# WHITE WOMEN
white_women_sum <- paste0(collapse = ' + ',
  '–29.799 * ln(age)', # Ln Age (y) –29.799
  
  '4.884 * ln(age)^2', # Ln Age, Squared 4.884
  
  '13.540 * ln(tc_value_as_number)', # Ln Total Cholesterol (mg/dL) 13.540

  '–3.114 * ln(age) * ln(tc_value_as_number)', # Ln Age × Ln Total Cholesterol –3.114
  
  '–13.578 * ln(hdl_value_as_number)', # Ln HDL-C (mg/dL) –13.578
  
  '3.149 * ln(age) * ln(hdl_value_as_number)', # Ln Age × Ln HDL-C 3.149
  
  'case when htt is true then
    2.019 * ln(sbp_value_as_number) else 0 end', # Ln Treated Systolic BP (mm Hg) 2.019
  
  NULL, ### # Ln Age × Ln Treated Systolic BP N/A
  
  'case when htt is false then
    1.957 * sbp_value_as_number else 0 end', # Ln Untreated Systolic BP (mm Hg) 1.957
  
  NULL, ### # Ln Age × Ln Untreated Systolic BP N/A
  
  '7.574 * smoker', # Current Smoker (1=Yes, 0=No) 7.574
  
  '–1.665 * ln(age) * smoker', # Ln Age × Current Smoker –1.665
  
  '0.661 * diabetes' # Diabetes (1=Yes, 0=No) 0.661
)

black_women_sum <- ''
white_men_sum <- ''
black_men_sum <- ''

individual_sum <- paste0(
  'case when is_women then (
      case when is_black then (', black_women_sum, ')
      else (', white_women_sum, ')
      end)
  else (
    case when is_black then (', black_men_sum, ')
    else (', white_men_sum, ')
    end)
  end')

```

## Results

## Plots

### Author contact  
Fabrício Kury  
`r format(Sys.time(), '%Y/%b/%d')`  
Be always welcome to reach me at fab@kury.dev.  
```{r, include = FALSE}
# DBI::dbDisconnect(dbcon)
```
