---
title: "Computing body mass index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{computing_bmi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r base, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# kury - credenciais.R contains the cred$pg credentials.
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/espinha/credenciais.R'))
```

Here we compute the body mass index (BMI) formula:  
  
`body_mass_index = body_weight / (body_height * body_height)`  
  
The formula assumes weight is in kilograms, and height in meters.  
Furthermore, this vignette assumes:  

 - You have a SQL server at `localhost`. This example uses PostgreSQL.
 - You have an OMOP Common Data Model v5.4 in schema `cdm_new_york3` in that server.
 - Body weight records are in table `MEASUREMENT`, with `measurement_concept_id` of `3025315`, `Loinc 29463-7 Body weight`.
 - Body height records are in table `MEASUREMENT`, with `measurement_concept_id` of `3036277`, `Loinc 8302-2 Body height`.
 
## Setup Phea  
First we connect to the server and call `setup_phea()`. Calling `setup_phea()` is optional, but it allows us to use
the convenience functions `slqt()` and `sql0()`.  
```{r setup}
library(phea)
suppressPackageStartupMessages(library(dplyr))

# Connect to SQL server.
dbcon <- DBI::dbConnect(
    RPostgres::Postgres(),
    host = 'localhost',
    port = 7654,
    dbname = 'fort',
    user = cred$pg$user,
    password = cred$pg$pass)

# Provide the connection to Phea so we can use the sqlt() and sql0() shorthands.
setup_phea(dbcon, 'cdm_new_york3')
```

## Create components
`make_component()` receives a lazy table (`dbplyr`/`dplyr` interface) and returns a component to use in 
`calculate_formula()`.  
We must specify which column (in the lazy table) is the timestamp, and which column is the patient ID. Notice that we
don't use quotes "".  
```{r components}
# Weight records
# Loinc 29463-7 Body weight, concept ID 3025315
weight <- sqlt(measurement) |>
  filter(measurement_concept_id == 3025315) |>
  make_component(
    .ts = measurement_datetime,
    .pid = person_id)

# Height records
# Loinc 8302-2 Body height, concept ID 3036277
height <- sqlt(measurement) |>
  filter(measurement_concept_id == 3036277) |>
  make_component(
    .ts = measurement_datetime,
    .pid = person_id)
```
`sqlt(measurement)` returns a reference (lazy table) to table `MEASUREMENT` in `cdm_new_york3`. `dplyr::filter()`
selects the desired records using a SQL `WHERE` statement.  

## Calculate the phenotype  
Each component goes inside a `list` in the call to `calculate_formula()`. The names of the list items become 
the names available to use in the formula. For example, if you want the `value_as_number` column from component
`height`, use `height_value_as_number`.  
  
Notice that, because the body height records in cdm_new_york3 are in centimeters, we use a separate formula to produce
`height_in_meters`.  
```{r phenotype}
# BMI phenotype
bmi <- calculate_formula(
  components = list(
    weight = weight,
    height = height),
  fml = list(
    height_in_meters = 'height_value_as_number / 100',
    bmi = 'weight_value_as_number / (height_in_meters * height_in_meters)'))

kable(head_shot(bmi, blind = TRUE))
```
The conversion from centimeters to meters could also be done using `mutate()` before creating the component, or using
parentheses in the `bmi` formula. Either approach would be equivalent.  

`head_shot()` is a convenience function to peek into the first rows of the lazy table.  
  
## Plot the phenotype result  
`phea_plot()` creates an interactive chart using the `plotly` library. It requires specifying the ID of one patient.
```{r plot, out.width="100%", out.height = 500}
# Plot BMI phenotype for patient whose pid = 2105
# We don't need to plot column height_value_as_number.
bmi |>
  select(-height_value_as_number) |>
  phea_plot(pid = 2105)
```
## Computing aggregates from a phenotype result  
The result `bmi` returned by `calculate_formula()` is a lazy table. We can manipulate it further to produce a few aggregates:  
```{r aggregates}
# See summary statistics
bmi |>
  group_by(pid) |>
  summarise(
    min_bmi = min(bmi, na.rm = TRUE),
    avg_bmi = mean(bmi, na.rm = TRUE),
    max_bmi = max(bmi, na.rm = TRUE),
    n_distinct_bmi = n_distinct(bmi)) |>
  arrange(desc(n_distinct_bmi), avg_bmi) |>
  head_shot(blind = TRUE) |>
  kable()
```

## Obtain SQL query that computes the phenotype  
To see the SQL query underlying the phenotype, use helper function `code_shot()`, or `dbplyr::sql_render()`, or the `.clip_sql` option in `calculate_formula()`.  
```{r, eval = FALSE}
code_shot(bmi)
```
```sql
`r dbplyr::sql_render(bmi)`
```

```{r, include = FALSE}
DBI::dbDisconnect(dbcon)
```
