---
title: "Getting started with Phea"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(credx) # contains fabcred$pg
# library(phea)
devtools::load_all()
suppressPackageStartupMessages(library(dplyr))

# Connect to SQL server.
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
  host = fabcred$pg$host, port = fabcred$pg$port, dbname = fabcred$pg$database,
  user = fabcred$pg$user, password = fabcred$pg$pass)

# Provide the connection to Phea so we can use the sqlt() and sql0() shorthands.
setup_phea(dbcon, 'cdm_new_york3')
```
## Background
Phea is a framework for computing electronic [patient] phenotypes:  

 - using formulas written in SQL language,  
 - and without using SQL joins.  
 
The intended use case is where you have a database with patient records and want to compute a formula. However, in
principle, Phea's approach need not be restricted to this use case.  

Phea passes formulas unmodified to the SQL query, so they need not be restricted to mathematical operators. The formulas 
can include any SQL expression that would be valid in a `SELECT` statement, such as `CASE WHEN ...` expressions and function
calls.  

To use Phea, you need a DBI connection to the SQL server, generated by functions such as `DBI::dbConnect()` or
`DatabaseConnector::connect()`.  

Phea also includes a function to create an interactive plot of the phenotype results in one patient. Below is an example of the body mass index formula.  
```{r, include=FALSE}
# Weight records
# Loinc 29463-7 Body weight, concept ID 3025315
weight <- sqlt(measurement) |>
  filter(measurement_concept_id == 3025315) |>
  make_component(
    .ts = measurement_datetime,
    .pid = person_id)

# Height records
# Loinc 8302-2 Body height, concept ID 3036277
height <- sqlt(measurement) |>
  filter(measurement_concept_id == 3036277) |>
  make_component(
    .ts = measurement_datetime,
    .pid = person_id)

bmi <- calculate_formula(
  components = list(
    weight = weight,
    height = height),
  fml = list(
    height_in_meters = 'height_value_as_number / 100',
    weight_in_kg = 'weight_value_as_number',
    body_mass_index = 'weight_in_kg / (height_in_meters * height_in_meters)'))
```
```{r bmi_plot, out.width="100%", out.height = 500, echo = FALSE}
bmi |>
  select(-weight_value_as_number, -height_value_as_number) |>
  phea_plot(pid = 2105, verbose = FALSE)
```
## Intuition
_Phea goes over patient records in chronological order._ Say you want to compute a formula:

`body_mass_index = body_weight / (body_height * body_height)`

Phea will compute `body_mass_index` for each person using the most recently available data at each point in
time.  

In this example, the formula depends on `body_weight` and `body_height`. Therefore, their timestamps will be the
timestamps for the resulting `body_mass_index`. This means that at every point in time where there is a `body_weight`
*or* `body_height` a record for a given patient, Phea will try to calculate `body_mass_index` again.  

Suppose a patient had a `body_weight = 75 kg` measurement on May 1st, a `body_height = 1.8 meters` on August 1st, then a
new `body_weight = 79 kg` on October 1st. Phea would produce 3 result rows as follows:

_May 1st:_  
`body_weight = 75 kg`  
`body_height = NULL`  
`body_mass_index = 75 / (NULL * NULL) = NULL`  

_August 1st:_  
`body_weight = 75 kg`  
`body_height = 1.8 meters`  
`body_mass_index = 75 / (1.8 * 1.8) = 23.14815`   

_October 1st:_  
`body_weight = 79 kg`  
`body_height = 1.8 meters`  
`body_mass_index = 79 / (1.8 * 1.8) = 24.38272`   

The algorithm continues as long as new records are avaialable. For example, if on November 1st a new record
`body_height = 1.78 meters` appears, the result would be:  

_November 1st:_  
`body_weight = 79 kg`  
`body_height = 1.78 meters`  
`body_mass_index = 75 / (1.78 * 1.78) = 24.93372`  

At each point in time, the most recently available data is used, unless you specify otherwise when creating a
component.  

## How to set up
First you need to connect to your relational server. **While Phea is not restricted to PostgreSQL, so far it has only been tested on PostgreSQL version 15.**  
```{r sql_connection}
# library(phea)
devtools::load_all()

# Connect to SQL server
# (fabcred credentials are not declared in this file)
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
  host = fabcred$pg$host, port = fabcred$pg$port, dbname = fabcred$pg$database,
  user = fabcred$pg$user, password = fabcred$pg$pass)
```

Calling `setup_phea()` is optional, but it enables the use of convenience functions `sql0()` and `sqlt()`.  

```{r phea_setup}
# Setup Phea and define cdm_new_york3 as the default schema.
setup_phea(
  connection = dbcon,
  schema = 'cdm_new_york3')
```

## How to compute a formula
Each formula computation comes from combining three parts:  

* One or more `record sources`  
* One or more `components`  
* One or more `formulas`  

### Record sources
A `record source` is any SQL query that provides a table with the data you want to use. For example, `select * from procedure_occurrence;`. It can be arbitrarily complex, with any number of clauses, subqueries, `JOIN`, `UNION`, etc. The `record source` is assumed to be a _long table_, that is each row corresponds to one record, and among the columns there is one that contains a patient identifier (`pid`) and one that contains a timestamp (`ts`).  

For example, suppose you have an OMOP Common Data Model schema called `cdm_new_york3`, and you would like to compute a
formula involving hemoglobin A1c levels.  

 - Such hemoglobin A1c records are in table `MEASUREMENT` inside schema `cdm_new_york3`.  
 - They are identified by the OMOP CDM concept ID `3004410`, corresponding to LOINC code `4548-4`, `Hemoglobin A1c/Hemoglobin.total in Blood`.  
 - The patient identifiers are in column `person_id`.  
 - The timestamps are in column `measurement_datetime`.  

That would constitute a record source. Let us create it in Phea:  

```{r record_source}
hemoglobin_query <- sql0('
  select * from cdm_new_york3.measurement
  where measurement_concept_id = 3004410
')

hemoglobin_record_source <- make_record_source(
  records = hemoglobin_query,
  ts = 'measurement_datetime',
  pid = 'person_id')
```

### Components
A `component` is a specification of _what records_ to pick from `record source`. You provide a `record source` to function `make_component()`, and it returns a `component`. If the records you want to use in the formula are simply the most recently available at each point in time, just use the default values:  

```{r component}
hemoglobin_component <- make_component(hemoglobin_record_source)
```

The `hemoglobin_component` can now be used to call `calculate_formula()`.  
  
The other records that can be picked by `make_component()` are:  
 - **`line = N` (a.k.a. `lag(N)`): pick the (N+1)th most recent record.** `line = 1` gives the 2nd most recent record, `line = 2` the third, etc. If you ask for the 2nd most recent (`line = 1`), but the patient only has 1 instance of that record, the value is NULL.  
 - **`delay = T`: pick the most recent record that is at least time distance `T` older than the phenotype date.** For example, if you put in a formula a diagnosis component with `delay('60 days')`, and a lab result component without `delay`, the result will include only the cases when the timestamp of the diagnosis is at least 60 days older than the timestamp of the lab result. If there is no such record, the value is NULL.  

## Calculate a formula  
For the sake of the example, let us suppose the phenotype that you want to compute is merely the square root of the patient's hemoglobin A1c level.  
  
`a1c_sqrt = sqrt(hemoglobin A1c level)`  
  
The numeric value of hemoglobin A1c is located in column `value_as_number` in the record source. The call to 
`calculate_formula()` is then:  
```{r formula}
phen <- calculate_formula(
  components = list(
    hemo_a1c = hemoglobin_component),
  fml = list(
    a1c_sqrt = 'sqrt(hemo_a1c_value_as_number)'))
```

Phea automatically recognizes that `hemo_a1c_value_as_number` means column `value_as_number` in the SQL table associated
with the component you named `hemo_a1c` in the `list()` when you called `calculate_formula()`. This is the syntax for accessing column values.    

Object `phen`, returned by `calculate_formula()`, is a lazy table in the `dplyr`/`dbplyr` framework. It can be materialized into a tibble by calling `dplyr::collect()`, or manipulated further using Phea or `dplyr`/`dbplyr`.

```{r collect_head}
dplyr::collect(head(phen)) |>
  knitr::kable()
```

To obtain just the SQL code, use the convenience function `code_shot()`, or the option `.clip_sql` in `calculate_formula()`, or `dbplyr::sql_render()`:  

```{r code_shot, eval = FALSE}
code_shot(phen)
```
```sql
`r code_shot(phen)`
```
Multiple formulas can be computed at once:  

```{r cascaded}
phen <- calculate_formula(
  components = list(
    hemo_a1c = hemoglobin_component),
  fml = list(
    a1c_sqrt = 'sqrt(hemo_a1c_value_as_number)',
    a1c_half_sqrt = 'a1c_sqrt / 2'))

dplyr::collect(head(phen)) |> knitr::kable()
```

## Example with more components  
So far the examples showed formulas with only one component. Suppose you would like to compute:  
`phen = sqrt([hemoglobin a1c]) * [patient's age at the time of measurement]`  

To obtain the patient's age, we can collect the date of birth from table `person`, then create a component from it.
```{r patient_age}
person_records <- make_record_source(
  records = sqlt(person),
  ts = 'birth_datetime',
  pid = 'person_id')

person_component <- make_component(person_records)

phen <- calculate_formula(
  components = list(
    hemo_a1c = hemoglobin_component,
    person = person_component),
  fml = list(
    pat_age = 'age(hemo_a1c_measurement_datetime, person_birth_datetime)',
    a1c_times_age = 'hemo_a1c_value_as_number * extract(year from pat_age)'))

dplyr::collect(head(phen)) |> knitr::kable()
```

Phea includes in the result all data points used in the formula computation. If other columns from the same rows are
wanted, `export = ` can be used:  
```{r export}
phen <- calculate_formula(
  components = list(
    hemo_a1c = hemoglobin_component,
    person = person_component),
  fml = list(
    pat_age = 'age(hemo_a1c_measurement_datetime, person_birth_datetime)',
    a1c_times_age = 'hemo_a1c_value_as_number * extract(year from pat_age)'),
  export = c(
    'hemo_a1c_provider_id',
    'hemo_a1c_visit_occurrence_id',
    'person_gender_concept_id'))

dplyr::collect(head(phen)) |> knitr::kable()
```


## Using records other than the one most recently available  
When creating a component, arguments `line` and `delay` can be used to pick records other than the one most recently
available.

`line` causes records to be _skipped_ in reverse chronological order. `line = 0` (default) gives the most recent
record available. `line = 1` skips 1 record, that is, gives the second most recent. `line = 2` skips 2 records, and so
on.
```{r line}
hemoglobin_component <- make_component(
  input_source = hemoglobin_record_source)

previous_hemoglobin_component <- make_component(
  input_source = hemoglobin_record_source,
  line = 1)

phen <- calculate_formula(
  components = list(
    hemo_a1c = hemoglobin_component,
    prev_hemo_a1c = previous_hemoglobin_component),
  fml = list(
    a1c_difference = 'hemo_a1c_value_as_number - prev_hemo_a1c_value_as_number'),
  export = c(
    'hemo_a1c_visit_occurrence_id',
    'prev_hemo_a1c_visit_occurrence_id'))

dplyr::collect(head(phen)) |> knitr::kable()
```

`delay` imposes a minimum time difference between the timestamp of the component and the timestamp of the phenotype.
It is defined in SQL language. For example, `delay = "'1 day'::interval` gives the most recently available record that
is at least 1 day older than the phenotype (in PostgreSQL). If there are no such records, the value is empty.
```{r delay}
hemoglobin_component <- make_component(
  input_source = hemoglobin_record_source)

old_hemoglobin_component <- make_component(
  input_source = hemoglobin_record_source,
  delay = "'6 years'::interval")

phen <- calculate_formula(
  components = list(
    hemo_a1c = hemoglobin_component,
    old_hemo_a1c = old_hemoglobin_component),
  fml = list(
    six_year_diff = 'hemo_a1c_value_as_number - old_hemo_a1c_value_as_number'),
  export = c(
    'hemo_a1c_visit_occurrence_id',
    'old_hemo_a1c_visit_occurrence_id'))

dplyr::collect(head(phen)) |> knitr::kable()
```

As of now, `line` and `delay` **cannot** be used simultaneously in the same component. This is a planned feature. If
both are provided, `line` is considered and `delay` is ignored. `delay` is only considered when `line = NA`.  
  
  
## Author contact
Contact Fabrício Kury at fab@kury.dev.  

```{r disconnect, echo = FALSE}
DBI::dbDisconnect(dbcon)
```


