---
title: "Find myocardial infarction in obese patients"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculate drop in body weight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
_This vignette assumes a SQL server at `localhost` (we use PostgreSQL), with data in OMOP Common Data Model v5.4 format in schema `cdm_new_york3`. The patient records shown in this example are synthetic data from [Synthea(TM) Patient Generator](https://github.com/synthetichealth/synthea)._  
```{r, results = 'hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
# kury - credenciais.R contains the cred$pg credentials.
source(paste0(Sys.getenv("HOME"), '/Estudo/Tecnologia/R/espinha/credenciais.R'))
```
```{r setup, print = FALSE}
library(phea)
library(dplyr)

# Connect to SQL server.
dbcon <- DBI::dbConnect(RPostgres::Postgres(),
  host = 'localhost', port = 7654, dbname = 'fort',
  user = cred$pg$user, password = cred$pg$pass)

# Call setup_phea so we can use sqlt() and sql0().
setup_phea(dbcon, 'cdm_new_york3')
```

How to find patients who had an acute myocardial infarction (AMI) when they were clinically obese (body mass index, BMI > 30)?  

A patient can have multiple myocardial infarctions, and their weight can change over time. In this example we are only interested in cases where the AMI happened _while_ the patient's BMI was _known_ to be > 30.  

The problem is, we cannot go back in time and measure the BMI of the patient on the day the AMI happened. Maybe someone did measure the patient's weight on that day, but maybe not. So, what do we say when we mean "known" to be obese?  

Maybe the only weight measurement available for that person is from 3 years before, or after, that day. Or maybe the AMI happened between two weight measurements, one giving BMI > 30 hence obesity, another giving BMI < 30 hence no obesity. Which one do you pick?  

Here, we will pick whichever combination of records is _closest in time to the AMI event_. Say a patient:  

 - was weighed by a nurse on day 1, and found to be obese,  
 - then the next available weight measurement on record is from day 120, and he was NOT obese anymore,  
 - and that patient had an AMI on day 40,  
 
...that means that on day 40, the closest weight record available was the one from day 1, which said that the patient was obese. So, we count that as a case of AMI happening in a patient known to be obese.  

That would not be the case, had the AMI been on day 90, for example. Similarly, if that same patient had a second AMI (god forbid) on day 120 or 121, that second AMI would count in the non-obese group.   

## Create components  
Let us first compute body mass index like in [another vignette](https://fabkury.github.io/phea/computing_bmi.html).  
```{r bmi_components}
# Weight component
# Loinc 29463-7 Body weight, OMOP concept ID 3025315
weight_component <- sqlt(measurement) |>
  filter(measurement_concept_id == 3025315) |>
  make_component(
    .ts = measurement_datetime,
    .pid = person_id)

# Height component
# Loinc 8302-2 Body height, OMOP concept ID 3036277
height_component <- sqlt(measurement) |>
  filter(measurement_concept_id == 3036277) |>
  make_component(
    .ts = measurement_datetime,
    .pid = person_id)
```

To which we add a component for acute myocardial infarction (AMI), from table `CONDITION_OCCURRENCE`.  

```{r ami_component}
# Acute myocardial infarction (AMI) component
# SNOMED 22298006 Myocardial infarction, OMOP concept ID 4329847
ami_component <- sqlt(condition_occurrence) |>
  filter(condition_concept_id == 4329847) |>
  make_component(
    .ts = condition_start_datetime,
    .pid = person_id)
```

## Calculate the phenotype  
Spelling out the BMI formula is straightforward:  

`bmi = height/(weight * weight)`

From which we can write a formula for "is obese":  

`is_obese = bmi > 30`

Notice that while the first formula is numerical (gives a number), the second one is logical (gives a `TRUE` or `FALSE`). Both are supported transparently, as long as they're not mixed, e.g. `3 / TRUE`.  

However, doing the same for "has an AMI" may not be as clear. What is the formula for "patient has an AMI event"?  

Since we have a column that gives the date when the AMI happened (it's called `condition_start_datetime`), we can check whether that column is empty. In each patient, for all phenotype calculations with a timestamp past the timestamp of an AMI event, that column will be populated by Phea with the date of that AMI event. For timestamps prior to the patient's first AMI, the column is empty. If the patient has a new AMI, the date is updated.   

To check if a column is empty in SQL, we use `is not null`:  

`has_ami = ami_condition_start_datetime is not null`

Armed with this, we can write the final formula for "has AMI and is obese":  

`obese_ami = is_obese AND has_ami`

```{r phenotype}
# AMI in obese patient
ami_obese <- calculate_formula(
  components = list(
    weight = weight_component,
    height = height_component,
    ami = ami_component),
  fml = list(
    
    height_in_meters = 'height_value_as_number / 100',
    
    bmi = 'weight_value_as_number / (height_in_meters * height_in_meters)',
    
    has_bmi = 'bmi is not null',
    
    is_obese = 'bmi > 30',
    
    has_ami = 'ami_condition_start_datetime is not null',
    
    ami_obese = 'has_ami and is_obese'
    ))
```
```{r filter ami}
# Keep only patients who had AMI
ami_patients <- ami_obese |>
  filter(has_ami) |>
  select(pid) |>
  distinct() |>
  pull()

ami_obese <- ami_obese |>
  filter(pid %in% local(ami_patients))

head_shot(ami_obese) |>
  kable()
```
## Keep only the records closest to each other  
Phea will compute `ami_obese` for all points in time in each patient. If a patient has a single AMI event, but had his body weight and height measured on 6 different dates, that is 7 computations of `ami_obese`, unless the date of the AMI coincides with one of the dates of the body measurements.  

Since we want to consider only the body measurements made closest in time to the AMI event, we can filter the rows by the `window`.  

Phea returns the `window` alongside each phenotype calculation. The `window` is the time distance between the oldest record, and the most recent one, that went into the phenotype calculation.  

Therefore, in each patient, for each AMI event, whichever row has the smallest `window`, that row is using the body measurements that are closest possible, in time, to the AMI event.  


```{r}
# Keep only the smallest window for each patient:
ami_obese_f <- ami_obese |>
  filter(has_ami & has_bmi) |>
  keep_row_by('window', c('pid', 'ami_condition_start_datetime'))

head_shot(ami_obese_f) |>
  kable()
```

## Plot the phenotype for a random patient
```{r}
ami_obese_patients <- ami_obese |>
  filter(ami_obese) |>
  select(pid) |>
  distinct() |>
  pull()

random_patient <- sample(ami_obese_patients, 1)
message('Sampled patient: ', random_patient)
```
```{r plot, out.width="100%", out.height = 1200}
ami_obese |>
  select(
    -height_value_as_number,
    -ami_condition_start_datetime) |>
  phea_plot(pid = random_patient)
```

## Obtain the SQL query that computes the phenotype  
To see the SQL query underlying the phenotype, use helper function `code_shot()`, or `dbplyr::sql_render()`, or the `.clip_sql` option in `calculate_formula()`.  
```{r, eval = FALSE}
code_shot(ami_obese)
```
```sql
`r code_shot(ami_obese)`
```

```{r, include = FALSE}
DBI::dbDisconnect(dbcon)
```
